\paperTitle

### 2014-02-10 doc/icml2014/directed.tex
Added proof for assumption 2
Updated pseudolikelihood and added something to the sample complexity section

Note that in each of the above cases,
  for any clique $\sC$, each hidden variable $h_i \in \sC$ has its own
  view which is conditionally independent given the other hidden variables in
  the clique.
%Fundamentally, algorithm takes advantage of having a view for every
  %hidden variable in a clique to help identify its hidden state.
We capture this general property as follows:
\begin{property}(Exclusive views)
  \label{prop:exclusive-views}
  A hidden clique $\sC$ of $\sG$ is said to possess the \textbf{exclusive views
  property} if for every hidden variable $h_i \in \sC$,
  there is some observed variable, $x_{v}$, which is conditionally
  independent of the rest of the clique given $h_i$,
  and whose conditional
  moment $\mOpp{v}{i} \eqdef \Pr(x_{v} \mid h_i)$ can be estimated consistently.
We call $x_v$ the exclusive view of $h_i$ in $\sC$.
\end{property}

@@

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnClique~(pseudoinverse)}
  \label{algo:learnclique}
  \begin{algorithmic}
    \REQUIRE Hidden clique $\sC$ with exclusive views (\propertyref{exclusive-views}).
    \ENSURE Marginal distribution of the clique $Z_\sC$.
      \STATE Identify exclusive views $\Sx{\sV} = \{x_{v_1}, \cdots, x_{v_m}\}$.
      \STATE Return $\mH_\sC \gets \mO_{\Sx{\sV}}( \pinv{\mOpp{v_1}{i_i}}, \cdots, \pinv{\mOpp{v_m}{i_m}} )$.
  \end{algorithmic}
\end{algorithm}
Of course, for $Z_{\sC}$ to be identifiable, we need various rank conditions,
which will be established in \sectionref{sampleComplexity}.

@@

\begin{lemma}[Canonical form]
  \label{lem:reduction}
Every directed graphical model can be transformed into one in which
  the observed variables are leaves with exactly one parent. 
There is a one-to-one correspondence between the parameters of this
  transformed model and the original one.
\end{lemma}
\begin{proof}
  \begin{figure}
    \centering
    \subimport{figures/}{reduction.tikz}
    \caption{Reduction to canonical form.}
    \label{fig:reduction}
  \end{figure}

@@

  Then, there is a one-to-one correspondence between every value of
  $\hp$ and $x_v$. Consequently, for any clique $\sC \contains \hp$, the
  parameters in the original graphical model can be obtained by
  substituting $\hp$ with $x_v$.
\end{proof}

\begin{lemma}[Bottlenecks guarantee exclusive views]
  Provided \propertyref{bottleneck} holds, every hidden variable
    $h_{i_1}, \cdots, h_{i_m}$ in a clique $\sC$ has a exclusive view.
\end{lemma}
\begin{proof}
  By \lemmaref{reduction}, we can assume that every observed
  variable is a leaf in the graph, causing any v-structures in the graph
  to be active. This allows us to reason about independence simply by
  checking whether or not two nodes are connected or not. Note that this
  proof applies for undirected graphs as well.

  Let $h$ be a hidden variable in $\sC$, let $\sV_{h}$ be the set of
    views for it and let $\sC^- \eqdef \sC \setminus \{h\}$.
  By definition, the views of $h$ are conditionally independent, viz.
    every connecting path between two views $x_1, x_2 \in \sV_h$ must pass through
    $h$.
  A view $x \in \sV_{h}$ is exclusive for $h$ iff $x \not\in \sV_{h'} ~ \forall h'
  \in \sC^-$.

  Suppose $h$ had no unique view, then $\sV_{h} \subseteq \Union_{h' \in \sC^-} \sV_{h'}$. 
  However, this would imply that for every $x_1, x_2 \in \sV_{h}$, there
    exists some $h_1, h_2 \in \sC^-$ (not necessarily distinct) such that
    $x_1$ is connected to $h_1$ and $x_2$ is connected $h_2$.
  By participation in the clique, either $h_1 = h_2$ or $h_1$ is
    connected to $h_2$, implying that $x_1$ and $x_2$ are connected
    via a path that does not pass through $h$, contradicting the
    statement that $x_1, x_2 \in \sV_h$.

@@

  Algorithmically, finding the unique view for $h$ can be done by
  subtracting the views of all other variables in the clique from
  $\sV_h$,i.e. $x_{v_h} \in \sV_h \setminus \Union_{h' \in \sC} \sV_{h'}$
\end{proof}

@@

To apply $\TensorFactorize$, we need that each $\mOpp{v}{i}$ has full
  column rank and that $\pi^{(i)} \eqdef \Pr(h_i) \succ 0$.
But $\mOpp{v}{i}$ is a product of
tensors on the path from $h_i$ to $x_v$, marginalizing out the rest of the graph.
To make this property more interpretable, let us assume the following property:%\verify
\begin{assumption} 
\label{asm:full-rank-plus}
For every clique $\sC \in \sG$ (including ones involving observed variables),
every $R^{k \times k}$ slice of the marginals $Z_\sC$ has full column rank $k$.
\end{assumption}
In our directed grid example (\sectionref{directedExample}), for the clique $\sC = \{ h_1, h_2 \}$ this
  condition implies $\pi = \Pr(h_1) \succ 0$ and $T = \Pr(h_2 \mid h_1)$ has column rank $k$; 
for the clique $\sC = \{ h_1, x_1 \}$, it implies $O = \Pr(x_1 \mid h_1)$ has column rank $k$

>>>>

Note that in each of the above cases,
  for any clique $\sC$, each hidden variable $h_i \in \sC$ has its own
  view which is conditionally independent given the other hidden variables in
  the clique.
%Fundamentally, algorithm takes advantage of having a view for every
  %hidden variable in a clique to help identify its hidden state.
We capture this general property as follows:
\begin{property}(Exclusive views)
  \label{prop:exclusive-views}
A hidden sub-graph $\sG' \subseteq \sG$ (typically a clique $\sC \in
  \sG$) is said to possess the \textbf{exclusive views property} if for
  every hidden variable $h_i \in \sG'$, there is some observed variable,
  $x_{v}$, which is conditionally independent of the rest of the clique
  given $h_i$, and whose conditional moment $\mOpp{v}{i} \eqdef
  \Pr(x_{v} \mid h_i)$ can be estimated consistently.
We call $x_v$ the exclusive view of $h_i$ in $\sG'$.
\end{property}

For the rest of this section, we will only consider sub-graphs which are
  cliques. 
In \sectionref{undirected}, we will show how the generalization
  to sub-graphs allows us to efficiently estimate the pseudo-likelihood.

% \begin{property}(Exclusive views)
%   \label{prop:exclusive-views}
%   A hidden clique $\sC$ of $\sG$ is said to possess the \textbf{exclusive views
%   property} if for every hidden variable $h_i \in \sC$,
%   there is some observed variable, $x_{v}$, which is conditionally
%   independent of the rest of the clique given $h_i$,
%   and whose conditional
%   moment $\mOpp{v}{i} \eqdef \Pr(x_{v} \mid h_i)$ can be estimated consistently.
% We call $x_v$ the exclusive view of $h_i$ in $\sC$.
% \end{property}

@@

\begin{algorithm}
  \caption{\LearnClique~(pseudoinverse)}
  \label{algo:learnclique}
  \begin{algorithmic}
    \REQUIRE Hidden clique $\sC$ with exclusive views (\propertyref{exclusive-views}).
    \ENSURE Marginal distribution of the clique $Z_\sC$.
      \STATE Identify exclusive views $\Sx{\sV} = \{x_{v_1}, \cdots, x_{v_m}\}$.
      \STATE Return $\mH_\sC \gets \mO_{\Sx{\sV}}( \pinv{\mOpp{v_1}{i_i}}, \cdots, \pinv{\mOpp{v_m}{i_m}} )$.
  \end{algorithmic}
\end{algorithm}
Of course, for $Z_{\sC}$ to be identifiable, we need various rank conditions,
which will be established in \sectionref{sampleComplexity}.

@@

\begin{lemma}[Canonical form]
  \label{lem:reduction}
Every directed (undirected) graphical model can be transformed into one in which
  the observed variables are leaves with exactly one parent (neighbour). 
There is a one-to-one correspondence between the parameters of this
  transformed model and the original one.
\end{lemma}
\begin{proof}
  \begin{figure}
    \centering
    \subimport{figures/}{reduction.tikz}
    \caption{Reduction to canonical form.}
    \label{fig:reduction}
  \end{figure}

@@

  Then, there is a one-to-one correspondence between every value of
  $\hp$ and $x_v$. Consequently, for any clique $\sC \contains \hp$, the
  parameters in the original graphical model can be obtained by
  substituting $\hp$ with $x_v$.

  This procedure applies straightforwardly for undirected graphical
  models, considering the neighbors $\sN(x_v)$ instead of its parents
  and children.
\end{proof}

\begin{lemma}[Bottlenecks guarantee exclusive views]
  \label{lem:bottleneck-views}  
Let $\sG' = \{ h_{i_1}, \cdots, h_{i_m} \} \subseteq \sG$ be
  a connected sub-graph of hidden variables. If every hidden variable
  $h_i \in \sG'$ is a bottleneck, then they each have an exclusive view.
\end{lemma}
\begin{proof}
  By \lemmaref{reduction}, we can assume that every observed
  variable is a leaf in the graph, causing any v-structures in the graph
  to be active. This allows us to reason about independence simply by
  checking whether or not two nodes are connected or not. Note that this
  proof applies for undirected graphs as well.

  Let $h$ be a hidden variable in $\sG'$, let $\sV_{h}$ be the set of
  views for it and let ${\sG'}^- \eqdef \sG' \setminus \{h\}$.
  By definition, the views of $h$ are conditionally independent, viz.
    every connecting path between two views $x_1, x_2 \in \sV_h$ must pass through
    $h$.
  A view $x \in \sV_{h}$ is exclusive for $h$ iff $x \not\in \sV_{h'} ~ \forall h'
  \in {\sG'}^-$.

  Suppose $h$ had no unique view, then $\sV_{h} \subseteq \Union_{h' \in {\sG'}^-} \sV_{h'}$. 
  However, this would imply that for every $x_1, x_2 \in \sV_{h}$, there
  exists some $h_1, h_2 \in {\sG'}^-$ (not necessarily distinct) such that
    $x_1$ is connected to $h_1$ and $x_2$ is connected $h_2$.
  By participation in the sub-graph, either $h_1 = h_2$ or $h_1$ is
    connected to $h_2$, implying that $x_1$ and $x_2$ are connected
    via a path that does not pass through $h$, contradicting the
    statement that $x_1, x_2 \in \sV_h$.

@@

  Algorithmically, finding the unique view for $h$ can be done by
  subtracting the views of all other variables in the sub-graph from
  $\sV_h$,i.e. $x_{v_h} \in \sV_h \setminus \Union_{h' \in \sC} \sV_{h'}$
\end{proof}

@@

To apply $\TensorFactorize$, we need that each $\mOpp{v}{i}$ has full
  column rank and that $\pi^{(i)} \eqdef \Pr(h_i) \succ 0$.
But $\mOpp{v}{i}$ is a product of
tensors on the path from $h_i$ to $x_v$, marginalizing out the rest of the graph.
To make this property more interpretable, let us assume the following property:%\verify
\begin{assumption} 
\label{asm:full-rank-plus}
For every clique $\sC \in \sG$ (including ones involving observed variables),
  every $\Re^{k \times k}$ slice of the marginals $Z_\sC$ has full column
  rank $k$ constructed by summing out the remaining indices.
\end{assumption}
In our directed grid example (\sectionref{directedExample}), for the clique $\sC = \{ h_1, h_2 \}$ this
  condition implies $\pi = \Pr(h_1) \succ 0$ and $T = \Pr(h_2 \mid h_1)$ has column rank $k$; 
for the clique $\sC = \{ h_1, x_1 \}$, it implies $O = \Pr(x_1 \mid h_1)$ has column rank $k$.

### 2014-02-10 doc/icml2014/undirected.tex
Updated pseudolikelihood and added something to the sample complexity section

\paragraph{Pseudolikelihood.}
Unlike for directed models, in the undirected case,
graphs with high treewidth present a challenge
for even fully-supervised maximum likelihood due to the intractability of the log-partition function.
This motivates looking at pseudolikelihood objective of \citep{besag75pseudo} for a computationally more
efficient solution.
The pseudolikelihood objective sums over the log-probability of each variable $a$ given its neighbors $\sN(a)$:
$$L_\text{pseudo}(\theta) \eqdef \E_{x \sim \sD}\left[\sum_{a \in X \cup H} \log p_\theta(a \mid x_{\sN(a)})\right].$$
In the fully-supervised setting, pseudolikelihood provides computationally efficient consistent estimates,
albeit at the cost of statistical efficiency.
Coincidentally, this is the same high-level motivation for using method of moments in the first place.

However, pseudolikelihood also replaces the log-partition function $A(\theta)$
with a sum over the variables: $A_a(\theta; x_{\sN(a)}) = \sum_a \exp(\theta^\top\phi(a))$,
which cannot be computed from just the hidden clique marginals $Z_\sC$,
since the neighbors $\sN(a)$ in general do not form a clique.
However, we adapt our earlier algorithm ($\LearnClique$) to estimate the marginals over all neighborhoods $Z_{\sN(a)}$.
While this method is only viable for graphs of low degree ($|\sN(a)|$ is small),
it can handle graphs with high treewidth.
For example, in the grid model, the degree is only 4.
Note that neither sample nor computational complexity depends on the treewidth.

@@

In general, we might only have exclusive views (\propertyref{exclusive-views}) for a subset of the cliques.
In this case, we can estimate the expected features $\mu_\sC$ for those cliques
  and use posterior regularization \citep{graca08em} or generalized expectation criteria \citep{mann08ge}
  to encourage $\E_{p_\theta}[\phi(\vx_\sC,\vh_\sC)]$ to match $\mu_\sC$.
The resulting objective functions are still non-convex, but we expect
  local optima to be alleviated.

>>>>

\paragraph{Pseudolikelihood.}
Unlike for directed models, in the undirected case,
graphs with high treewidth present a challenge
for even fully-supervised maximum likelihood due to the intractability of the log-partition function.
This motivates looking at pseudolikelihood objective of \citep{besag75pseudo} for a computationally more
efficient solution.
The pseudolikelihood objective sums over the log-probability of each variable $a$ given its neighbors $\sN(a)$:
$$L_\text{pseudo}(\theta) \eqdef \E_{x \sim \sD}\left[\sum_{a \in X \cup H} \log p_\theta(a \mid x_{\sN(a)})\right].$$
In the fully-supervised setting, it is well-known that pseudolikelihood provides computationally efficient consistent estimates,
albeit at the cost of statistical efficiency.
Coincidentally, this is the same high-level motivation for using method of moments in the first place.

However, pseudolikelihood also replaces the log-partition function $A(\theta)$
with a sum over the variables in the neighbourhood: $A_a(\theta; x_{\sN(a)}) = \sum_{a' \in \{a\} \union \sN(a)} \exp(\theta^\top\phi(a'))$,
which cannot be computed from just the hidden clique marginals $Z_\sC$,
since the neighbors $\sN(a)$ in general do not form a clique.
We adapt our earlier algorithm ($\LearnClique$) to estimate the
  marginals over all neighborhoods $Z_{\sN(a)}$ by noting that
  \lemmaref{bottleneck-views} guarantees exclusive views for any connected
  sub-graph of $\sG$, not just cliques.

While this method is only viable for graphs of low degree, i.e.\@ when $|\sN(a)|$ is small,
it can handle graphs with high treewidth.
For example, in the grid model, the degree is only 4.
Note that neither sample nor computational complexity depends on the treewidth.

@@

In general, it might be case that we only have exclusive views (\propertyref{exclusive-views}) for a subset of the cliques.
In this case, we can estimate the expected features $\mu_\sC$ for those cliques
  and use posterior regularization \citep{graca08em} or the generalized expectation criteria \citep{mann08ge}
  to encourage $\E_{p_\theta}[\phi(\vx_\sC,\vh_\sC)]$ to match $\mu_\sC$.
The resulting objective functions are still non-convex, but we expect
  local optima to be alleviated.
